### 2018-05-25 latex/sc18/algorithm.tex
minor fixes
added table and some discussion in analysis section

We will analyze the cost of a single outer iteration.
While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.

\paragraph{\emph{Computation}}
The local computation occurs at \cref{line:locMTTKRP,line:hadamard,line:locNLS,line:locSYRK}.
The cost of \cref{line:hadamard} is $O(NR^2)$, the cost of \cref{line:locNLS} is $O(R^3I_n/P)$, which is a loose upper bound for BPP and other methods \cite{KBP16}, and the cost of \cref{line:locSYRK} is $O(R(I_n/P)^2)$.
The sum of these three costs across all inner iterations is $O(R^2N^2+(R^3/P+R/P^2)\sum I_n)$, which is dominated by the cost of the MTTKRP.
When using dimension trees to perform the MTTKRP (\cref{line:locMTTKRP}), we compute the cost amortized over all inner iterations.
In this case, the cost is dominated by the two partial MTTKRP computations (from the root of the tree), which together are $O((R/P) \prod I_n)=O(IR/P)$ and dominate the costs of the multi-TTVs.
We note that this cost involves the product of all the tensor dimensions, which is why it dominates, and we note that it scales linearly with $P$.

\paragraph{\emph{Communication}}
The communication within the inner iteration occurs at \cref{line:reduce-scatter,line:all-reduce,line:all-gather}.
\Cref{line:all-reduce} involves $O(R^2)$ data and a collective across all processors.
\Cref{line:reduce-scatter,line:all-gather} involve $O(I_nR/P_n)$ data across a subset of $P/P_n$ processors.
Thus, the All-Reduce dominates the latency cost and the Reduce-Scatter/All-Gather dominate the bandwidth cost, for a total outer iteration communication cost of $O(R\sum I_n/P_n)$ words and $O(N\log P)$ messages.
If the optimal processor grid can be chosen to minimize communication (assuming $P$ is sufficiently factorable), then the bandwidth cost can achieve a value of $O(RI^{1/N}/P^{1/N})$ by making the local tensors as cubical as possible.
We note that this cost scales with $P^{1/N}$, which is far from linear scaling.
However, it is proportional to the geometric mean of the tensor dimensions (on the order of one tensor dimension), which is much less than the computation cost dependence on the product of all dimensions.

\paragraph{\emph{Memory}}
The algorithm requires extra local memory to run.
Aside from the memory required to store the local tensor of $O(I/P)$ words and factor matrices of cumulative size $O((R/P)\sum I_n)$, each processor must be able to store a redundant subset of the rows of the factor matrices it needs to perform MTTKRP computations.
This corresponds to storing $P/P_n$ redundant copies of every factor matrix, which results in a local memory requirement of $O(R \sum I_n/P_n)$ for a general processor grid.
The processor grid that minimizes communication also minimizes local memory, and the extra memory requirement can be as low as $O(RI^{1/N}/P^{1/N})$.

The dimension tree algorithm also requires extra temporary memory space, but the space required tends to be much smaller than what is required to store the local tensor.
If the tensor dimensions can be partitioned into two parts with approximately equal geometric means, the extra memory requirement for running a dimension tree is as small as $O(R\sqrt{I}/P)$, which is typically dominated by $O(I/P)$.

>>>>

\begin{table*}
\centering
\begin{tabular}{|c|cccc|}
\hline
 & \textbf{Computation} & \textbf{Memory - Local MTTKRP} & \textbf{Communication} & \textbf{Memory - Par.~Algorithm}  \\
\hline
Par-NNCP w/ opt.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$& $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$  \\
Par-NNCP w/ gen.~proc.~grid & & & $R\sum_n \frac{I_n}{P_n}$ & $R\sum_n \frac{I_n}{P_n}$  \\
NoDimTree & $\frac{NIR}{P}$ & $\displaystyle \max_n \frac{RI/I_n}{P/P_n}$  &  & \\
\hline
\end{tabular}
\smallskip
\caption{Leading-order per-iteration costs in terms of computation (flops), communication (words moved), and memory (words).  We ignore constants but omit big-Oh notation for clarity.  The first line corresponds to \cref{alg:Par-NNCP-short} with an optimal choice of processor grid and applying the dimension tree optimization locally.  The second line corresponds to the changes in communication and memory for a general processor grid.  The third line corresponds to the changes in computation and memory if the dimension tree optimization is not applied.}
\label{tab:costs}
\end{table*}

We will analyze the cost of a single outer iteration.
While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.
We summarize the analysis in \cref{tab:costs}, showing the differences with and without an optimal processor grid and with and without using a dimension tree.

\paragraph{\emph{Computation}}
The local computation occurs at \cref{line:locMTTKRP,line:hadamard,line:locNLS,line:locSYRK}.
The cost of \cref{line:hadamard} is $O(NR^2)$, the cost of \cref{line:locNLS} is $O(R^3I_n/P)$, which is a loose upper bound for BPP and other methods \cite{KBP16}, and the cost of \cref{line:locSYRK} is $O(R(I_n/P)^2)$.
The sum of these three costs across all inner iterations is $O(R^2N^2+(R^3/P)\sum I_n+(R/P^2)\sum I_n^2)$, which is dominated by the cost of the MTTKRP.
When using dimension trees to perform the MTTKRP (\cref{line:locMTTKRP}), we compute the cost amortized over all inner iterations.
In this case, the cost is dominated by the two partial MTTKRP computations (from the root of the tree), which together are $O((R/P) \prod I_n)=O(IR/P)$ and dominate the costs of the multi-TTVs.
We note that this cost involves the product of all the tensor dimensions, which is why it dominates, and we note that it scales linearly with $P$.

\paragraph{\emph{Communication}}
The communication within the inner iteration occurs at \cref{line:reduce-scatter,line:all-reduce,line:all-gather}.
\Cref{line:all-reduce} involves $O(R^2)$ data and a collective across all processors.
\Cref{line:reduce-scatter,line:all-gather} involve $O(I_nR/P_n)$ data across a subset of $P/P_n$ processors.
Thus, the All-Reduce dominates the latency cost and the Reduce-Scatter/All-Gather dominate the bandwidth cost, for a total outer iteration communication cost of $O(R\sum I_n/P_n)$ words and $O(N\log P)$ messages.
If the optimal processor grid can be chosen to minimize communication (assuming $P$ is sufficiently factorable), then the bandwidth cost can achieve a value of $O(NRI^{1/N}/P^{1/N})$ by making the local tensors as cubical as possible.
We note that this cost scales with $P^{1/N}$, which is far from linear scaling.
However, it is proportional to the geometric mean of the tensor dimensions (on the order of one tensor dimension), which is much less than the computation cost dependence on the product of all dimensions.

\paragraph{\emph{Memory}}
The algorithm requires extra local memory to run.
Aside from the memory required to store the local tensor of $O(I/P)$ words and factor matrices of cumulative size $O((R/P)\sum I_n)$, each processor must be able to store a redundant subset of the rows of the factor matrices it needs to perform MTTKRP computations.
This corresponds to storing $P/P_n$ redundant copies of every factor matrix, which results in a local memory requirement of $O(R \sum I_n/P_n)$ for a general processor grid.
The processor grid that minimizes communication also minimizes local memory, and the extra memory requirement can be as low as $O(NRI^{1/N}/P^{1/N})$, which is typically dominated by $O(I/P)$.

The dimension tree algorithm also requires extra temporary memory space, but the space required tends to be much smaller than what is required to store the local tensor.
If the tensor dimensions can be partitioned into two parts with approximately equal geometric means, the extra memory requirement for running a dimension tree is as small as $O(R\sqrt{I/P})$, which is also typically dominated by $O(I/P)$.
